\section{Cosmology}
\subsection{Questions}
\begin{enumerate}
\item Put on a timeline, and describe the principal events in the thermal history of the
universe, from $kT = 10$ TeV to $kT = 0.1$ eV.
\item Give a semi-quantitative discussion of the connection between flucuations of the
cosmic microwave background on angular scales of arcminutes to degrees, and the
baryonic structures (galaxies, clusters, correlations of galaxies) observed in the local
universe, redshift $z < 0.5$.
\item Which elements/isotopes are produced in Big Bang Nucleosynthesis and in what
quantities? Explain qualitatively how the yield of each depends on the cosmic baryon
density and why.
\end{enumerate}

\subsection{Measuring Distances}
The basic concept of distance is the distance ladder, which is basically 
different techniques used to measure distances farther and farther away.  
There is overlap between the different techniques, which allows the next 
technique to be calibrated by the previous one.  The only problem with this 
is errors and uncertainties at the bottom of the ladder propogate all the 
way up.  

The first step is parallaxes, which are reliable out to $\sim100$ pc or so.  As 
the Earth moves around the Sun, the apparent positions of stars will change 
slightly relative to more distant fixed background stars.  The distance 
to an object in parsecs is defined to be $\frac{1}{p}$ where $p$ is the 
parallax in arcseconds.  One useful trick to know for quickly converting 
between angular size and physical size- the physical size in AU of something 
$d$ parsecs away with an angular size of $\theta$ arcsec is just $\theta d$.  
So, for example, if a star and planet are $1$ pc away and are separated by 
$1$ arcsec, the planet is $1$ AU from the star.  Parallax is the only 
distance indicator that is not model dependent or reliant on some other 
calibration.  

Out to distances of $\sim1$ kpc, main sequence fitting can be used to find 
distances to main sequence stars.  If the spectral type of a star is measured, 
and the absolute magnitude of that spectral type is known from a closer 
star with a parallax distance, the difference between apparent and absolute 
magnitude gives the distance to the new star.  

This has been done for clusters of stars that contain Cepheid variables, 
leading to the next step in the distance ladder.  Cepheids are horizontal 
branch high mass stars in the instability strip of the HR diagram (see Stars 
notes).  Cepheids are useflu as distance indicators because their periods 
are tightly correlated with their luminosities by 
\begin{equation}
\log{\frac{L}{L_{\odot}}}\sim2.4+1.2\log{\frac{P}{days}}
\end{equation}
So by measuring the period of pulsation, the intrinsic luminosity can be 
determined, giving a distance.  Nearby galaxies (within $20$ Mpc) are close 
enough that individual Cepheids can be observed and used to get distances.  

Advantages:  Bright ($M_V\sim-4$ to $-7$), so the can be used in other galaxies.  
Also, we understand the physics behind their pulsation well, and can identify 
them based on their period.  

Disadvantages:  Cepheids are relatively rare since they correspond to short 
phases of stellar evolution and high masses.  Also, there are uncertainites 
about systematics of the period-luminosity relation like how it depends on 
metallicity and color.  

The tip of the red giant branch method is applicable out to similar distances 
as Cepheids.  Stars at the top of the red giant branch, right before He flash, 
have a known absolute magnitude of $-4$.  This point also corresponds to a 
drop-off in the frequency of stars, so once this is found and its brightness 
is measured, the distance can be determined.  

Advantages: Stars are bright and method is relatively precise and 
well-understood.  

Disadvantages:  Only works for old populations.  

Another method is planetary nebulae luminosity functions.  Planetary nebulae 
have an observed cut-off in luminosity at absolute magnitude $-4.5$.  The 
physics behind the cut-off is understood, and planetary nebulae can be found 
in all types of galaxies.  

For galaxies out to $\sim100$ Mpc, there are several methods to get distances, 
including the Tully-Fisher relation, globular cluster  
luminosity functions, and surface brightness fluctuations.  The Tully-Fisher 
relation is an observed correlation between the rotational velocity of 
spiral galaxies and their luminosity:
\begin{equation}
L\propto v_{rot}^4
\end{equation}
The rotational velocity is usually measured by HI $21$ cm line width.  The 
problems with this method are the relation has $\sim10-20$\% scatter, and 
the physics behind it are not well understood.  The fundamental plane gives 
similar relations for elliptical galaxies.  Globular clusters have well-defined empirical luminosity functions that can be used as 
standard candels.  Globular clusters have a peak at an absolute magnitude 
of $-6.6$.  An advantage of this method is it is not affected by dust.  
However, deep photometry is needed to measure the luminosity function, and 
this method is not as precise as others.  Surface brightness fluctuations 
involve measuring the surface brightness (flux per solid angle) in several 
regions of the same galaxy.  The observed surface brightness depends on the 
number of stars in the region, which should be similar in the different 
patches.  If the average number of stars is N, the Poisson scatter in the 
number of stars (and thus the brightness) will be $\frac{1}{\sqrt{N}}$.  
The number of stars per solid angle will also depend on distance as 
$N\propto r^2$.  So the observed scatter $\sigma\propto\frac{1}{r}$.  This 
relation can be calibrated using nearby galaxies.   

For even more distant galaxies (out to $\sim1$ Gpc), Type Ia supernovae 
can be used as standard candles to get a distance.  A tight correlation 
exists between the shape of a Ia lightcurve and the luminosity of the 
supernova, so by measuring the shape of the light curve, the distance can 
be determined.  This method is very precise, but we still don't understand 
the mechanism for a type Ia supernova.  Type Ia supernovae can be observed 
into the Hubble flow, so a plot of distance versus redshift can be used 
to determine the cosmological parameters.  The main problem with all these 
methods so far is they rely on calibrations from other methods.  One of the 
main goals of the Hubble Space Telescope was to improve these calibrations 
and determine $H_0$ more precisely.  The main focus was to better calibrate 
Cepheids and the distance to the LMC and SMC, which other methods are based 
on.  The HST $H_0$ Key Project determined $H_0=72\pm3\pm7$ km/s/Mpc.

Two methods that don't need calibration, but instead are model dependent 
are the S-Z effect (see section in Radiative) and gravitational lense 
time delays.  If a background quasar is lensed, variability in the quasar 
will be seen at different times on different sides of the lense, since the 
path length is different.  If the mass distribution of the lense is known, 
then the ratio of the path length difference to the source distance and lense 
distance can be determined.  Since the time delay is known, the path length 
difference can be calculated, giving the distance to the lense and the 
background quasar.  The modeling is very complicated however.  

  
