\section{Stars}

FYI- I looked through the old qual questions we got, and it seems like almost everyone is asked 
something about the HR Diagram, so we should know that inside and out.  Also, questions about 
spectral lines are more popular than I thought they would be.  Globular clusters were also popular.

\subsection{Questions}
\begin{enumerate}
\item \textbf{Explain what is the Hayashi track, and describe what types of objects live on it.
      Qualitatively explain how it arises and what assumptions are required for its derivation.}

\item \textbf{Estimate the temperature as a function of depth in the Sun's convection zone. What is the temperature at the base of the convection zone?}

	%Imagine a blob of gas inside a gaseous medium whose density deviates from the density of its environment. If the blob's density is smaller than the surrounding density, the system is unstable to convection (we should also know how to express this in terms of entropy).
	%Imagine that the blob has $\rho_{b1},~P_{b1},~T_{b1}$, and the surrounding medium has $\rho_1,~P_1,~T_1$. The blob moves up a distance $\delta r$, where the environment has $\rho_2,~P_2,~T_2$, and now the blob has $\rho_{b2},~P_{b2},~T_{b2}$. 
	Assume the process is adiabatic, so the blob obeys $P = K \rho ^\gamma$, and assume we have an ideal gas, so $P = \frac{\rho k_{\rm B} T}{\mu m_p}$. First differentiate the ideal gas law to obtain
	
	\begin{equation}
	\frac{dP}{dr} = \frac{P}{\rho} \frac{d \rho}{dr} + \frac{P}{T} \frac{d T}{dr} \,\, ,
	\end{equation}
	assuming that the pressure does not vary much with the mean molecular weight $\mu$. We can also differentiate the adiabatic pressure equation
	\begin{equation}
	\frac{dP}{dr} = \gamma \frac{P}{\rho}\frac{d\rho}{dr}\,\, .
	\end{equation}
	Setting these two equal, we get
	\begin{equation}
	\biggl(\frac{dT}{dr}\biggr)_{\rm ad} = \biggl(1 - \frac{1}{\gamma}\biggr)\frac{T}{P} \frac{dP}{dr}\,\, ,
	\end{equation}
	which is the adiabatic temperature gradient. In a star, we can also use the ideal gas law to replace $P$ and hydrostatic equilibrium to replace $dP/dr$:
	\begin{equation}
	\biggl(\frac{dT}{dr}\biggr)_{\rm ad} = -\biggl(1 - \frac{1}{\gamma}\biggr)\frac{\mu m_p}{k_{\rm B}} \frac{G M_r}{r^2}\,\, .
	\end{equation}
	This describes how the temperature in the bubble changes as it rises and expands adiabatically. If $dT/dr$ is steeper than the adiabatic temperature gradient, the fluid is superadiabatic and there will be convection.
	
	The convection zone of the Sun extends about $R_\astrosun/3$ down from the surface. Assume $M_r$ is constant and is the mass of the Sun, i.e. relatively little mass is contained in the convection zone. For an ideal monatomic gas, $\gamma = 5/3$, and let's assume $\mu \approx 1/2$, since in the outer layers of the Sun it should be mostly hydrogen. We can integrate our equation to get $T(R)$:
	
	\begin{equation}
	\int^{T(R)}_{T_{\rm eff}} dT = -\frac{m_p}{5 k_{\rm B}} G {\rm M}_\odot \int^R_{\rm R_\odot} r^{-2}dr \,\, .
	\end{equation}
	Alternatively, you could integrate from the base of the convection zone if you know what the temperature is there ($1.8 \times 10^6~K$) and can remember that it's about a third of the way in. The final formula is:
	\begin{equation}
	T(R) = \frac{m_p}{5 k_{\rm B}} G {\rm M}_\odot \biggl( \frac{1}{R} - \frac{1}{{\rm R}_\odot}\biggr) + 5800~{\rm K}\,\, .
	\end{equation}
	And you can plug numbers in yourself (someone check the form of this equation and make sure io got it right).
	
\item \textbf{Describe the major sources of opacity in stars, and how each depends on density,
      temperature, and metallicity.}
      
      \newthought{There roughly four} regimes that are important for determining the
      continuum opacity of the star.  In order of increasing temperature: blending of
      spectral lines, H$^-$ absorption, neutral H absorption, and electron scattering.
      Let's discuss each of these in turn.

      \newthought{At very low temperatures,} (ie. in the atmospheres of M dwarfs and brown dwarfs),
      where $T<3000$ K (I'm guessing at this temperature -- Michael), the number of molecular
      and atomic lines can become overwhelming.  At a certain point, the lines begin to fall
      on top of each other and act as a source of continuum opacity.  At a certain point, the
      opacity becomes so great that it influences the structure of star\sidenote{
        For instance in brown
        dwarfs, the blended line opacity is so great that light escapes the surface of the star
        only in regions where there aren't very many molecular lines.  In the infrared,
        color filters are designed on the same criteria (we don't want atmospheric lines
        contaminating our images).  Therefore, brown dwarfs have almost all of their light
        coming out in bands centered on the color filters Y, J, H, and K.  As a side-effect,
        this also makes brown dwarfs look very blue.
      }.

      Electron scattering (Thomson): This opacity is due to photons scattering off free electrons. This opacity is generally close to constant, since the Thomson cross-section is a constant given by
      \begin{equation}
      \sigma_{\rm T} = \frac{8 \pi e^4}{3 c^4 m_e^2} = 6.6524\times 10^{-25}~{\rm cm}^2\,\, .
      \end{equation}
      The opacity $\kappa = \sigma / m$, so the opacity will depend on the total mass per scattering electron; that is, it will depend on the composition of the material and also how much it is ionized. Because of the ionization state, it will also depend on the temperature, i.e. whether the material is fully ionized. For a fully ionized medium, however, the opacity should be roughly constant, since there will typically be 1 or 2 baryons per electron, so $m \sim 1-2~m_p$. Therefore $\kappa$ is usually of the order 0.1-0.4 cm$^2$ g$^{-1}$.

      The exact opacity (assuming a fully ionized gas) from Thomson scattering can be calculated 
      as a function of H mass fraction.  In a similar calculation to the mean molecular weight 
      derivation, 
      \begin{equation}
      n_e=n_H+2n_{He}+\Sigma{A}{2}n_A=\frac{\rho}{m_H}\left(X+\frac{2}{4}Y+\frac{1}{2}Z\right)=\frac{\rho}{2m_H}\left(1+X\right)
      \end{equation}
      So,
      \begin{equation}
      \kappa_T=\frac{n_e\sigma_T}{\rho}=\frac{sigma_T}{2m_H}\left(1+X\right)=\left(1+X\right)0.2~{\rm cm^2 g^-1}
      \end{equation}
       
      H$^-$ ion: Occasionally, if there are enough free electrons present (easier with increased partial ionization of heavy elements), some will become weakly bound to hydrogen atoms, so that the proton is surrounded by two electrons. Opacity is caused by photoionization of this extra electron with binding energy of 0.75 eV. 
      
      \begin{equation}
      \kappa_{H^{-}} \propto Z \rho^{1/2} T^{9}
      \end{equation}
      
      The opacity is very sensitive to temperature, because a fragile balancing act is required. The temperature must be less than 10,000K (when Hydrogen ionizes) so that neutral Hydrogen is around to get another electron. But, the temperature must also be high enough so that an abundance of free electrons are around which were liberated from ionized heavy elements. This opacity is also sensitive to metallicity, for the same reason. 
      Despite $H^{-}$ being a fragile and rare ion, this type of opacity is the most important source in our Sun's atmosphere, and is dominant in FGK stars.

      Neutral H Transitions:
      These include bound-bound transitions and bound-free transitions.  In bound-bound 
      transitions, a photon is absorbed by an atom, moving an electron to a higher energy level.  
      These transitions have corresponding wavelengths asscociated with them, giving rise 
      to spectral lines.  Higher density leads to higher opacity, but the temperature 
      dependence of this opacity is complicated.  The temperature and composition determine 
      how many atoms of a certain type there are, and what state these atoms are in.  This 
      determines the strengths of spectral lines.  Of course, if the star is hot enough to 
      ionize everything, neutral transitions will not take place.
      
      Bound-free opactiy is the ionization of an atom (H is most important, since it's most 
      abundant).  For photons with enough energy to ionize H, the cross section depends varies 
      as $\lambda^3$.  On our stars midterm, we had to plot bound-free opacity as a function of 
      wavelength.  I can't draw it here, but it's good to know what the plot looks like.  
      Bound-free opacity is responsible for the Balmer jump is stellar spectra at 365 nm, where 
      H in the $n=2$ level can be ionized.  This produces a drop in flux from the star at shorter 
      wavelengths.

      Averaging bound-free and free-free absorption over all wavelenthts gives Kramer's opacity:
      \begin{equation}
      \bar{\kappa}\propto \frac{\rho}{T^{3.5}}
      \end{equation}
       
\end{enumerate}

\subsection{Stellar Properties}
Let's start by listing the properties of the Sun, which are just good to know.
\begin{dgroup}
\begin{dmath}
L_\astrosun = 3.8 \times 10^{33}~{\rm erg}~{\rm s}^-1
\end{dmath}
\begin{dmath}
R_\astrosun = 6.9 \times 10^{10}~{\rm cm}
\end{dmath}
\begin{dmath}
M_\astrosun = 2 \times 10^{33}~{\rm g}
\end{dmath}
\begin{dmath}
t_\astrosun = 4.5 \times 10^9~{\rm yrs}
\end{dmath}
\begin{dmath}
T_{\rm eff} = 5800~{\rm K}
\end{dmath}
\begin{dmath}
T_{\rm c} = 15 \times 10^6~{\rm K}
\end{dmath}
\begin{dmath}
\rho_{\rm c} = 150~{\rm g}~{\rm cm}^{-3}
\end{dmath}
\begin{dmath}
\langle \rho \rangle_\astrosun = 1~{\rm g}~{\rm cm}^{-3}
\end{dmath}
\end{dgroup}
Ranges for other stars:
\begin{dgroup}
\begin{dmath}
M \sim 0.1 - 100~{\rm M}_\astrosun
\end{dmath}
\begin{dmath}
R \sim 0.1 - 1000~{\rm R}_\astrosun
\end{dmath}
\begin{dmath}
L \sim 10^{-4} - 10^6~{\rm L}_\astrosun
\end{dmath}
\begin{dmath}
T_{\rm eff} \sim 1000 - 5 \times 10^4~{\rm K}
\end{dmath}
\end{dgroup}

\subsection{Spectral Types}
O, B, A, F, G, K, M\newline
Within these types, stars are numbered 0 through 9 (hottest to coolest).  Luminosity classes 
are also used.  I-IV are giant stars, while V are main sequence stars.  D is for white dwarfs.  
So, for example, the Sun is a G2V star.\newline
O stars:\newline
>16 $M_{\odot}$, $M_V=-5$ or brighter\newline
Very weak spectral lines (just about everything is ionized, so there's nothing 
to produce transitions).  Very weak Balmer lines and some weak lines to due 
ionization of He.  No Balmer break visible.\newline
B stars:\newline
2-16 $M_{\odot}$\newline
Stronger Balmer lines, also He absorption lines.  Still a relatively clean 
spectrum though.  Balmer break visible (and visible in all later spectral 
types.\newline
A stars:\newline
1.5-2 $M_{\odot}$, $M_V\sim 0$\newline
Strongest Balmer lines because these are the hottest stars (hot enough to have 
lots of H in the n=2 state) without being too hot and ionizing all the H.  
Effective temperature of $\sim$10,000 K is perfect for Balmer lines.\newline
F stars:\newline
1-1.5 $M_{\odot}$\newline
Balmer lines start to weaken again.  Star seeing metal lines (mainly neutral 
or singly ionized Ca, Mg, and Na).\newline
G stars:\newline
0.8-1 $M_{\odot}$, $M_V\sim4$\newline
This includes the Sun.  Weaker Balmer lines, stronger metal lines.\newline
K stars:\newline
0.5-0.8 $M_{\odot}$\newline
Weaker Balmer, stronger metal.  Also start seeing absorption bands from 
molecules (smearing of lots of absorption lines due to rotational, vibrational, 
and electronic transitions).
M stars:\newline
<0.5 $M_{\odot}$, $M_V\sim 8-12$\newline
Dominated by molecular bands, in particular TiO between 6600\AA\ and 8600\AA.

\subsection{HR Diagram}
Be able to draw and explain one.  Some useful approximate relations for the main sequence:
\begin{equation}L\propto T_{eff}^8\end{equation}
\begin{equation}R\propto T_{eff}^2\end{equation}
\begin{equation}L\propto M^3\ for\ stars\ more\ massive\ than\ the\ Sun\end{equation}
\begin{equation}L\propto M^5\ for\ stars\ less\ massive\ than\ the\ Sun\end{equation}
Main sequence masses range from $\sim 0.1\ M_{\odot}$ to $\sim 100\ M_{\odot}$.  Main sequence 
radii range from $\sim 0.25\ R_{\odot}$ to $\sim 25\ R_{\odot}$.

\subsection{Equations of Stellar Structure}
\newthought{There are three} key pieces of physics that are necessary for understanding stars:
\begin{enumerate}
    \item Force balance: pressure vs. gravity (and sometimes rotation)
    \item Energy transport: conduction, radiation, convection
    \item Energy generation: fusion, gravitational contraction
\end{enumerate}
These form the essence of the equations of stellar structure.
\begin{dgroup*}
\begin{dmath}
    \frac{\d M_r}{\d r} = 4\pi r^2\rho \quad \text{(mass continuity)}
\end{dmath}
\begin{dmath}
    \frac{\d P}{\d r} = -\frac{GM_r\rho}{r^2} \quad \text{(hydrostatic equilibrium)}
\end{dmath}
\end{dgroup*}

Let's talk first about force balance. Assuming no radiation, this is just a competition between pressure and gravity. We can look at a layer in the star with pressure $P$ at its base and $P+dP$ above, and we know the force on the layer due to gravity. The total force will be the pressure difference times the area of the shell $A$ (which will eventually drop out) plus the gravitational force $-\frac{G M_r M_{\rm shell}}{r^2}$. If we assume hydrostatic equilibrium, the total force is zero. We then can switch to differentials to get the hydrostatic equilibrium equation:

\begin{equation}
\frac{dP}{dr} = -\frac{\rho G M_r}{r^2}\,\,.
\end{equation}

The mass continuity equatin just comes from the amount of mass in a spherical shell at radius r 
and of thickness dr.

The radiative temperature gradient can be derived by considering the energy in shells at different 
radii.  Since energy flows out of the star, the energy density must decrease outwards.  Consider 
a shell with energy denstity $u+du$ (shell 1), with another shell $dr$ above it of energy 
density $u$ (shell 2).  The energy flow from shell 1 to shell 2 per unit time ($L(r)$) is 
the excess energy in shell 1 divided by the time it takes for this energy to flow from shell 1 to 
shell 2.  So
\begin{equation}\label{eq:L}
L(r)=-\frac{4\pi r^2\,dr\,du}{(dr)^2/lc}
\end{equation}
where the denominator is the time it takes a the photon to cross from shell 1 to shell 2 by random 
walk.  Taking the angle of the radiation into account introduces a factor of 1/3 on the right hand 
side.  Equation \ref{eq:L} can be rewritten as 
\begin{equation}
\frac{L(r)}{4\pi r^2}=-\frac{cl}{3}\frac{du}{dr}
\end{equation}
This can also be thought of as a diffusoin equation, where the left hand side is energy flux, 
$\frac{cl}{3}$ is the diffusion coefficient, and $\frac{du}{dr}$ is the energy density gradient.  
Plugging in $u=aT^4$ and $l=\frac{1}{\kappa\rho}$, and solving for $\frac{dT}{dr}$ gives
\begin{equation}
\boxed{\frac{dT(r)}{dr}=-\frac{3L(r)\kappa(r)\rho(r)}{4\pi r^2 4acT^3(r)}}
\end{equation}

The energy conservation equation just comes from defining $\epsilon(r)$ as the power generated 
per unit mass.  With this definition,
\begin{equation}
\boxed{\frac{dL(r)}{dr}=4\pi r^2\rho (r)\epsilon (r)}
\end{equation}

In addition to the four main equations of stellar structure, three more equations must be 
specified for the pressure, opacity, and energy generation.  
\begin{equation}
P=P(\rho,T,composition)
\end{equation}
\begin{equation}
\kappa=\kappa(\rho,T,composition)
\end{equation}
\begin{equation}
\epsilon=\epsilon(\rho,T,composition)
\end{equation}
Also, the four differential equations require four boundary conditions:
\begin{equation}
M(r=0)=0
\end{equation}
\begin{equation}
L(r=0)=0
\end{equation}
\begin{equation}
P(r=R)=0
\end{equation}
\begin{equation}
M(r=R)=M_*
\end{equation}

So, there are 7 equations for 7 unknowns, and 4 boundary conditions for 4 differential equations.  
This means that there is a unique solution given the composition and total mass of the star, the 
only two free parameters in the equations.  This is the \emph{Vogt-Russell conjecture}, that the 
properties and evolution of a star depend only on its mass and initial composition.

The equation of state is almost always ideal gas, but here's a good derivation of mean molecular 
weight, which I personally can never remember.

For a fully ionized gas, H contributes 2 particles, He contributes 3, and metals contribue 
$\sim\frac{A}{2}$.  So the total number density of particles is
\begin{equation}
n=2n_H+3n_{He}+\Sigma \frac{A}{2}n_A=\frac{rho}{m_H}\left(2X+\frac{3}{4}Y+\frac{1}{2}Z\right)=\frac{\rho}{2m_H}\left(3X+\frac{1}{2}Y+1\right)
\end{equation}
using the fact that $X+Y+Z=1$.  Then
\begin{equation}
\mu=\frac{\rho}{nm_H}=\frac{2}{1+3X+0.5Y}
\end{equation}

io likes the following definition of mean molecular weight:
\begin{equation}
\frac{1}{\mu_I} = \sum \frac{X_i}{A_i}\,\, ,
\end{equation}
\begin{equation}
\frac{1}{\mu_e} = \sum \frac{Z_iX_i}{A_i} \,\, ,
\end{equation}
\begin{equation}
\frac{1}{\mu} = \biggl(\frac{1}{\mu_I} + \frac{1}{\mu_e} \biggr)^{-1}\,\,.
\end{equation}

See the will ask question for a discussion of opacity.

Also, radiation pressure can contribute to the overall pressure (and sometimes dominate).  
For a photon gas,
\begin{equation}
P_{rad}=\frac{1}{3}u=\frac{1}{3}aT^4
\end{equation}

\subsection{The Hayashi Track}
\newthought{The Hayashi track}
describes the evolution of star that is entirely convective
(with a narrow radiative envelope for the photosphere).
Hayashi derived these tracks by showing that for
a given mass, there is an effective temperature below which no stable solutions exist.
That is, any star with an effective temperature below the limit will collapse until its
effective temperature is above the limit.  In this way, stars essentially slide along
the boundary of stability, maintaining a constant effective temperature
(even as their central temperature increases).

Let's see a brief derivation of this.  To begin with we assume that the interior is entirely
convective so that the pressure is given by the adiabatic relationship
\begin{dmath}
    P\sim\rho^\gamma
\end{dmath}.
By combining this with the ideal gas law, we find that
\begin{dmath}\label{eq:pt_adiabatic}
    P\sim T^{\frac{\gamma}{\gamma-1}}
\end{dmath}.
For a monatomic gas, $\gamma=5/3$, so that $P\sim T^{5/2}$.

Next we can use the equation of hydrostatic equilibrium to estimate the central pressure $P_c$ and
temperature $T_c$ as\sidenote{
    To see this, approximate $\d P/\d r$ as $P_c/R$ and approximate $\rho$ as $M/R^3$.
    Forget about numerical constants because we're just looking for a scaling relationship
    here.  The temperature comes from the ideal gas law.
}:
\begin{dgroup*}
\begin{dmath*}
    P_c \sim \frac{M^2}{R^4}
\end{dmath*}
\begin{dmath*}
    T_c \sim \frac{M}{R}
\end{dmath*}
\end{dgroup*}
With these expressions for the central pressure and temperature, we can use
Equation~\ref{eq:pt_adiabatic} to find the pressure and temperature anywhere in the
star -- notably at the surface of the photosphere.

However, before we can do that, we need to locate the photosphere.  To do this, we
assume an opacity law of the form $\kappa \sim \rho T^a$, and that the photosphere
can be treated in the plane-parallel approximation\sidenote{
    This assumes that the photosphere is narrow and contains a negligible amount of mass.
}.
In this case the equation of hydrostatic equilibrium becomes
\begin{dmath*}
\frac{\d P}{\d z} = -\rho g
\end{dmath*},
where $g = GM/R^2$.  The photosphere is located where $\tau\sim 1$ (or $\tau\sim 2/3$ if
you're being picky), so let's replace $\d z$ with the optical depth $\d\tau = -\kappa\rho\d z$
such that
\begin{dmath*}
\frac{\d P}{\d\tau} = \frac{g}{\kappa}
\end{dmath*}.
Integrating to the surface of the photosphere therefore gives us
\begin{dmath}
    P_{\rm eff} \sim \frac{g}{\kappa} \nolinebreak
                \sim \frac{R}{T_{\rm eff}^a}
\end{dmath}.
Using Equation~\ref{eq:pt_adiabatic} and our expressions for $P_c$, $T_c$, and $P_{\rm eff}$
we can solve for $T_{\rm eff}$ to find (taking $\gamma=5/3$)
\begin{dmath}\boxed{
    T_{\rm eff} \sim R^{2.5/(2.5+a)}M^{0.5/(2.5+a)}
}\end{dmath}.
Therefore when the opacity is very temperature sensitive\sidenote{
    For H$^-$ opacity, $a\sim10$.
}, the effective temperature is almost independent of radius.  Therefore the Hayashi
track is a vertical line on the HR diagram.

It is worth noting here that it is critical for the opacity to increase rapidly with temperature.
As the star contracts, the central temperature of the star increases.  In fact, at any fixed
position within the star, the temperature increases.  However, if the opacity is very temperature
sensitive, the photosphere will stay at a location where the temperature remains roughly
constant.  This means that the effective temperature stays constant.

\subsection{The Henyey Track}

\newthought{In the core}
of a protostar, Kramer's opacity law is a good description of the mean opacity.
As the protostar contracts along the Hayashi track, the temperature of the core increases
even as the effective temperature remains constant.  The opacity in the core therefore decreases
as the star contracts and eventually the core becomes radiative.  As the star continues to
contract, the opacity in the core further decreases and the radiative region grows in size.
Therefore the structure of the star responds to further contraction by lowering the opacity
in its core and moving more energy through radiation.  This allows the star to push more energy
out (it's more transparent to radiation!), and hence the luminosity increases.  Because the
luminosity is increasing while the radius is shrinking, the effective temperature must also
increase (it can no longer remain constant as on the Hayashi track)\sidenote{
    Recall that $L=4\pi R^2\sigma T_{\rm eff}^4$ to good approximation.
}.  Therefore, when the core becomes radiative, the protostar moves to higher luminosities and
higher effective temperatures (after moving to lower luminosities with a constant effective
temperature on the Hayashi track).  This track is occasionally called the Henyey track.

\subsection{Scaling Relations on the Main Sequence}
Assume that P(r), M(r), $\rho (r)$, and T(r) are roughly power laws.  Then from the equations of stellar structure we have:

\begin{equation}
P \thicksim \frac{\rho M}{r}
\end{equation}
\begin{equation}
M \thicksim r^3\rho
\end{equation}
\begin{equation}
L \thicksim \frac{T^4r}{\kappa \rho}
\end{equation}

For moderately massive stars greater than $1 M_\odot$, the pressure is dominated by kinetic gas pressure and the opacity is dominated by electron scattering.

\begin{equation}
P \thicksim \rho T
\end{equation}
\begin{equation}
\kappa = constant
\end{equation}

Equating hydrostatic equilibrium with the equation of state:

\begin{equation}
P \thicksim \frac{M\rho}{r} \thicksim \rho T
\end{equation}
\begin{equation}
T \thicksim \frac{M}{r}
\end{equation}

Replacing T in the luminosity equation yeidls:
\begin{equation}
L \thicksim M^3
\end{equation}

Furthermore, $M \thicksim R$ on the main sequence, which implies a constant core temperature.  To see this, consider a star that is contracting and heating up.  An equilibrium will be established when the density and the temperature in the core are high enough for the onset of nuclear reactions.  Since the nuclear power density $\epsilon$ depends mainly on temperature, for any initial mass, the radius of the star will stop shrinking when a particular core temperature is reached.  Therefore the internal temperature is comparable in all main sequence stars.  From stellar models, over a range of around 100 in mass, the core temperature varies only by a factor of 4.  

For stars less massive than the sun, the opacity is dominated by bound-bound and bound-free opacities, which follow kramer's opacity $\kappa \thicksim \rho T^{-3.5}$.  Since the temperature is approximately constant:

\begin{equation}
 M \thicksim R
\end{equation}
\begin{equation}
\kappa \thicksim \rho \thicksim \frac{M}{r^3} \thicksim M^{-2}
\end{equation}

The luminosity equation becomes:

\begin{equation}
L \thicksim \frac{M}{M^{-2}M^{-2}} \thicksim M^{5}
\end{equation}

For the most massive stars, radiation pressure dominates.  The equation of state then goes as $P \thicksim T^4$, and electron scattering opacity dominates.  The mass-luminosity relationship then goes as:

\begin{equation}
L \thicksim M
\end{equation}





\subsection{Stellar Evolution}

Collapse from a gas cloud:

The first important thing to know is the Jeans criterion. You can get here by considering a deviation from a system in hydrostatic equilibrium, which is described by the Virial Theorem:

\begin{equation}
2K + U = 0\,\, ,
\end{equation}
where we can approximate the kinetic and gravitational potential energies as

\begin{equation}
U \sim \frac{3 G M_{\rm c}^2}{5 R_{\rm c}}
\end{equation}
\begin{equation}
K = \frac{3}{2} N k_{\rm B} T = \frac{3 M_{\rm c} k_\rm B T}{\mu m_{\rm H}}
\end{equation}

You can assume a constant density for the cloud and find the minimum mass necessary to start collapse, $M_{\rm c} > M_{\rm J}$:

\begin{equation}
M_{\rm J} \approx \biggl( \frac{5 k_{\rm B}T}{G \mu m_{\rm H}}\biggr)^{3/2}\biggl( \frac{3}{4\pi\rho_0}\biggr)^{1/2}
\end{equation}

We can also express this criterion as $R_{\rm c} > R_{\rm J}$, where

\begin{equation}
R_{\rm J} \approx \biggl( \frac{15 k_{\rm B} T}{4 pi G \mu m_{\rm H} \rho_0} \biggr)^{1/2}
\end{equation}

Once the Jeans criterion is satisfied, the cloud basically collapses in freefall at the beginning. It is isothermal, optically thin, and radiates energy efficiently. The freefall time is

\begin{equation}
t_{\rm ff} = \biggl( \frac{3 \pi}{32 G \rho_0}\biggr)^{1/2} \sim \biggl( \frac{1}{G \rho_0}\biggr)^{1/2}
\end{equation}
The numerical factor is from the derivation in Carrol \& Ostlie; you can derive this by considering a particle on the outer edge of and object falling inward due to the gravitational force of all the interior mass.

Also note that fragmentation is an issue: that is, the Jeans mass changes as the density of the cloud goes up, and a natural consequence of this is that a smaller mass is required to collapse. Therefore, one large cloud (of, say, 50 solar masses) might collapse into $\sim 50$ solar-mass clouds.

What eventually stops fragmentation is that the cloud stops being isothermal. The cloud starts transporting more energy adiabatically rather than radiatively because it starts to become optically thick, and radiation does not transport energy out efficiently enough.

(There is some stuff about shocks produced by the outer layers of the cloud free-falling onto the nearly hydrostatic star, which powers it for a while. Not sure how much we need to know about this.)

As a fragmented clump collapses, it must have a way of releasing the energy from the collapse, or 
the temperature and pressure will rise and stop the collapse.  At this stage, the excess energy 
goes into first dissociating H$_2$, then ionizing atomic H.  Once (almost) all the H has been 
ionized, the star must release its gravitational energy by radiating.  This slows the collapse, 
and a slowly contracting proto-star forms, with contraction governed by the Kelvin-Helmholtz 
timescale.  

Pre-Main-Sequence Evolution:

First, the \textbf{pre-MS protostar} collapses on the Kelvin-Helmholtz timescale $t_{\rm KH}$, which for $1~{\rm M}_\astrosun$ is about $10^7$ years. The Kelvin-Helmoholtz timescale is given by
\begin{equation}
t_{\rm KH} \approx \frac{GM^2}{RL}\,\, .
\end{equation}
This just comes from approximating the energy available for the luminosity is the gravitational potential energy, so the timescale for radiating this energy is just the total gravitational potential energy divided by the luminosity.

The effective temperature increases, so the opacity of the outer layers becomes dominated by the H$^-$ ion, which gets its extra electrons from increased partial ionization of heavy elements. This increased opacity causes the envelope of the protostar to become convective and sometimes this envelope can become very large, reaching even to the center of the star. This object evolves along the Hayashi track, its luminosity decreasing while the effective temperature increases only slightly (it is almost constant, $\sim 3800$ K).

A \textbf{solar mass star on the Hayashi track} is completely convective and starts burning deuterium in the first million years, which slows the collapse slightly and for a short time, since deuterium is not very abundant. The central temperature rises, and opacity in the center decreases, so the core becomes radiative. This causes the luminosity to start increasing. At the same time, the core becomes hot enough to start more nuclear reactions (dominated by the first two steps of the PP chain and the CNO cycle. Nuclear energy generation starts dominating over luminosity from gravitational collapse. A large temperature gradient develops in the core, which starts to become partly convective again, and then the core expands a bit, which decreases the luminosity because $\epsilon_{\rm grav}$ becomes negative, and $\epsilon = \epsilon_{\rm nuc} + \epsilon_{\rm grav}$. When C-12 is exhausted, the core adjusts and PP-I chain dominates, and the star settles onto the main sequence.

For \textbf{lower-mass stars}, the temperature might never get hot enough to burn C-12, so they don't turn up as much in luminosity, but rather fall closer to the Hayashi track down onto the Main Sequence. Below about 0.08 M$_\odot$, no nuclear reactions can be sustained, and these stars never reach the Main Sequence. Lower mass starts also never develop radiative cores, so they are always fully convective.

\textbf{Massive stars} start burning carbon and hydrogen much more quickly because the central temperature rises faster. These break away from the Hayashi track earlier and at higher luminosities, since nuclear reactions contribute more to the luminosity earlier on. The CNO cycle becomes the dominant energy generation process because of the high temperatures. As the star travels horizontally on the HR diagram, the core remains convective because the energy generation is so temperature-dependent, while the envelope is radiative.


\subsection{Radial Velocities}

Derivation of radial velocity amplitude from a star and a low mass companion (say a planet).

Let $r_p$ be the distance between the planet and the center of mass of the system, and $r_s$ be the distance between the star and the center of mass.

\begin{equation}
a = r_p + r_s
\end{equation}

\begin{equation}
v_p = \frac{2\pi r_p}{P}
\end{equation}
\begin{equation}
v_s = \frac{2\pi r_s}{P}
\end{equation}

\begin{equation}
M_pr_p = M_sr_s
\end{equation}

Combining above equations yields the following relation:
\begin{equation}
\frac{v_p}{v_s} = \frac{r_p}{r_s} = \frac{M_s}{M_p}
\end{equation} 

\begin{equation}
a = r_p + r_s = (\frac{M_s}{M_p} + 1)r_s = r_s\big(\frac{M_s + M_p}{M_s}\big)
\end{equation}

Plugging the above expression into Kepler's 3rd law:

\begin{equation}
P^2 = \frac{4\pi^2r_s^3(M_p + M_s)^2}{GM_p^3}
\end{equation}

Plugging in for $r_s$ and rearranging a bit:

\begin{equation}
v_s^3 = \frac{P^{-1}2\pi GM_p^3}{(M_p + M_s)}
\end{equation}

By assuming the companion is significantly less massive than the star, you can simplify the expression for the radial velocity amplitude K.  
\begin{equation}
v_s = K = 2\pi G P^{-\frac{1}{3}}M_pM_s^{-\frac{2}{3}}
\end{equation}

\subsection{Atmospheres}
The Boltzmann and Saha equations are probably more involved than we need to know (especially the 
Saha equation), but here they are anyway.
\begin{equation}
\frac{n_1}{n_2}=\frac{g_1}{g_2}e^{-(E_1-E_2)/kT}
\end{equation}
\begin{equation}
\frac{n_{i+1}}{n_i}=\frac{2Z_{i+1}}{n_eZ_i}\left(\frac{2\pi m_ekT}{h^2}\right)^{3/2}e^{-\chi_i/kT}
\end{equation}
where $n_{i+1}$ and $n_i$ are are the numbers of atoms in the $i$ or $i+1$ ionization state, 
$Z_i$ is the partition function of the $ith$ ionization state, and $\chi_i$ is the ionization 
energy.  All we should probably know for these equations relating to stars is that the describe 
the population of atoms in the star's atmosphere, and thus determine the strengths of spectral 
lines.  For example, the competing temperature dependence of these equations gives that the 
peak fraction of H atoms in the $n=2$ state happens at around 10,000 K.  This is why A stars show 
such strong Balmer lines.

\subsection{Interiors}


